A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-08 text/manuscript.Rmd

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: hide
    theme: flatly
    smart: true
bibliography: references.bib
---

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1g**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
distinguished  by KRAS mutations among others. But even PAAD comprises tumors
from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
replacing genes with pathways (**figure 2**). A complete list of pathway
disruptions with percent mutated samples and top genes is provided in
**supplementary table ST1**). These reveal the broad importance of many well
known pathways that are disrupted in cancer, including <mark>PI3K/Akt
activation, DNA repair etc fix this with top 5 or so quoted pathway
labels</mark>. The top third of pathways drop off gradually in overall mutation
frequency, possibly owing to relative diversity among all tumors.

![](../figures/figure2_temp.png)

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
which each column represents a tumor and each row represents a pathway. The
pathways are ranked such that the pathway with the most disruptions across all
7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
top pathway disruptions successively. All 377 pathways are shown. Red signifies
pathway disrupted pathways, grey signifies not disrupted.</p>

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
0 pathways in class 8 (for which most pathways are putatively disrupted), 25
pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
had enrichment. To explore the specific pathways marking each class, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
metabolic pathways including RNA and protein biosynthesis (**supplementary
figure S4**). Most class 4 enriched pathways were not unique to this class but
shared in class 10 & 12, suggesting that this grouping reflects either greater
heterogeneity or possibly a lower stage of tumor evolution (as distinct from
clinical ``stage''). Classes 9-12 have in common mutations in extracellular
signaling, intracellular signaling, and immune-related pathways (see **figures
3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
distinct of the 12, we chose to investigate these two groups further.

![](../figures/fig2B_short.png)

@@

## Supplementary Information

![**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._](../figures/figS2.png)

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_](../figures/figS3.png)

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_](../figures/figS4.png)
![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: hide
    theme: flatly
    smart: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1h**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
distinguished  by KRAS mutations among others. But even PAAD comprises tumors
from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
replacing genes with pathways (**figure 2**). A complete list of pathway
disruptions with percent mutated samples and top genes is provided in
**supplementary table ST1**). These reveal the broad importance of many well
known pathways that are disrupted in cancer, including <mark>PI3K/Akt
activation, DNA repair etc fix this with top 5 or so quoted pathway
labels</mark>. The top third of pathways drop off gradually in overall mutation
frequency, possibly owing to relative diversity among all tumors.
<div style="text-align: right"> Figure 2 Data </div>
```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

######################
## oncoprint top100 ##
######################
onco_color <- c("0" = "grey", "1" = "firebrick3")

top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class

top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths

top100path[top100path == "1"] <- "MUT" ## categorize boolean values
top100path[top100path == "0"] <- " "
  col <- c("MUT" = "firebrick3")
heatmap_legend <- list(title = "", at = c("MUT"),
                       labels = c("Mutation"))
alter_fun = list(
    background = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
            gp = gpar(fill = "#CCCCCC", col = NA))
    },
    # bug red
    MUT = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
            gp = gpar(fill = col["MUT"], col = NA))
    })
  column_title <- "Top 100 Pathway PanCancer"
  onco_res <- oncoPrint(top100path,
    alter_fun = alter_fun, col = col,
    column_title = column_title, heatmap_legend_param = heatmap_legend,
    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
    top_annotation = NULL
    )

clust_col_order <- names(top100path[,onco_res@column_order])
path_row_order <- rownames(top100path[onco_res@row_order,])

onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
  rownames_to_column(var = "path_name") %>% ## path_name to col
  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
save(onco_ggplot, file = "onco_ggplot.Rda")

#####################
## oncoprint panel ##
#####################
clust_col_order <- list()
path_row_order <- list()

for(n in 1:12) {
  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
  top10path[top10path == "0"] <- " "
  col <- c("MUT" = "firebrick3")
heatmap_legend <- list(title = "", at = c("MUT"),
                       labels = c("Mutation"))
alter_fun = list(
    background = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
            gp = gpar(fill = "#CCCCCC", col = NA))
    },
    # bug red
    MUT = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
            gp = gpar(fill = col["MUT"], col = NA))
    })
  column_title <- "test"
  onco_res <- oncoPrint(top10path,
    alter_fun = alter_fun, col = col,
    column_title = column_title, heatmap_legend_param = heatmap_legend,
    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
    top_annotation = NULL
    )
  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
}

save(clust_col_order, path_row_order, file = "onco_clust.Rda")
```

```{r figure-2-plot, fig.width=72, fig.height=48}
## top100 pathways pancancer oncoprint
load("../data/onco_ggplot.Rda") ## load data
onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint

fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
  geom_tile(aes(fill = as.character(mutation_status))) +
  scale_fill_manual(values = onco_color) +
  xlab(label = "sample_id") +
  labs(tag = "A") +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none")

## top10 pathways for each cluster
load("../data/onco_clust.Rda") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
figure_labs <- LETTERS[seq(from = 2, to = 13)]
for(n in 1:12) {
  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
  rownames_to_column(var = "path_name") %>% ## path_name to col
  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
  geom_tile(aes(fill = as.character(mutation_status))) +
  scale_fill_manual(values = onco_color) +
  xlab(label = paste0("Class ", n)) +
  labs(tag = figure_labs[n]) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none") 
}

## setting patchwork panel layout
layout <- '
AABCD
AABCD
AABCD
AAEFG
AAEFG
AAEFG
AAHIJ
AAHIJ
AAHIJ
AAKLM
AAKLM
AAKLM
'

fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
which each column represents a tumor and each row represents a pathway. The
pathways are ranked such that the pathway with the most disruptions across all
7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
top pathway disruptions successively. All 377 pathways are shown. Red signifies
pathway disrupted pathways, grey signifies not disrupted.</p>

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
0 pathways in class 8 (for which most pathways are putatively disrupted), 25
pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
had enrichment. To explore the specific pathways marking each class, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
metabolic pathways including RNA and protein biosynthesis (**supplementary
figure S4**). Most class 4 enriched pathways were not unique to this class but
shared in class 10 & 12, suggesting that this grouping reflects either greater
heterogeneity or possibly a lower stage of tumor evolution (as distinct from
clinical ``stage''). Classes 9-12 have in common mutations in extracellular
signaling, intracellular signaling, and immune-related pathways (see **figures
3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
distinct of the 12, we chose to investigate these two groups further.
<div style="text-align: right"> Figure 3 data </div>
```{r group-pathway-enrichment-data-generation, eval=FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns

## Getting the total number of tumor samples in each cluster
n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
  mutate(clust = as.character(clust)) ## convert factor to character

## Organizing rBeta data
load("../data/gdc_reactome_path_analysis.Rda")
clust_group <- n_tumor_in_clust$clust
cluster_path_count <- list()
for(n in clust_group) {
  print(n)
  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
    .[,"V1"] ## get sample_ids
  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
    'colnames<-' (n) ## change colname to cluster number
  cluster_path_count[[n]] <- path_mutation
  rm(path_mutation, sample_id, n)
  gc()
}
cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together

global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
  'colnames<-' (clust_group) %>% ## rename column with tumor class
  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
  gather(., key = "clust", value = "global_mut", -path)

rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples

rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
gc()

enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
                                            clust = rbeta_path_analysis$clust,
                                            fg_mut = rbeta_path_analysis$clust_mut,
                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
                                            bg_mut = rbeta_path_analysis$global_mut,
                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
                                            )
fg_mut_rate <- list()
bg_mut_rate <- list()
for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
  print(n)
  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
  fg_numerator <- sum(df$fg_mut)
  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
  bg_numerator <- sum(df$bg_mut)
  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
  rm(df)
}

fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
  rownames_to_column(var = "clust") ## move rownames to col
bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
  rownames_to_column(var = "clust") ## move rownames to col

enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
  left_join(., bg_mut_rate, by = "clust") %>%
  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
gc()

moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
})

moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
})

## ModerateEnrichment probability 
prob <- moderate_fg > moderate_bg ## boolean data
mean_prob <- apply(prob, MARGIN = 2, function(x) {
  mean(x)
})
rm(prob)
gc()
## Moderate Difference
CI <- 0.025 ## setting a 95% confidence interval

diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
rm(moderate_fg, moderate_bg, prob)
gc()

sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
}))
quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
  quantile(x, c(CI, 0.5, 1-CI))
}))

moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
                                   clust = enrich_prob_uncertainity_in_bg$clust,
                                   probability = mean_prob,
                                   signif = sig_fg_bg[,1],
                                   difference = quant_diff[,2],
                                   lower = quant_diff[,1],
                                   upper = quant_diff[,3])
rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
gc()
#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
load("../data/moderate_enrich_prob.Rda")

diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")

effect_size <- pheatmap(diff_heatmap,
         show_rownames = F,
         show_colnames = T,
         cluster_cols = T,
         cluster_rows = T,
         cellwidth = 10,
         cellheight = 0.76,
         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
         clustering_method = "ward.D2")

effect_row_order <- effect_size$tree_row$order
effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap

effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
  gather(., key = "cluster", value = "effect_size", -path_name)

effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))

effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
))
effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))

save(effect_heatmap, file = "effect_heatmap.Rda")
```

```{r group-pathway-enrichment heatmap}
load("../data/effect_heatmap.Rda") ## load data

effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme

fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
  geom_tile(size = 0.6, aes(fill = discrete_color)) +
  scale_fill_manual(values = effect_colors) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "none")

fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
  geom_bar(stat = "identity", width = 1) +
  theme_classic() +
  scale_fill_manual(values = effect_colors) + 
  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line = element_blank())

## setting patchwork panel layout
layout <- '
AAAAA
AAAAA
AAAAA
AAAAA
AAAAA
AAAAA
AAAAA
AAAAA
#BBB#
'

fig3 + fig3_legend + plot_layout(design = layout)
```

@@

## Supplementary Information
```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data

umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
  geom_point(size = 1) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
![**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]

```{r supplementary-figure-s2-data-generation}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
<div style="text-align: right"> Supplementary Figure S3 Data </div>
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
gc()
gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1

##oncogene
oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
##tumor_supressor
tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor

selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data

selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join

selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_mut, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d"), meausure.vars = tumor_suppressor$external_gene_name) %>%
  filter(value != 0)

save(selected_mut, file = "selected_mut.Rda")
```

```{r supplementary-figure-s3}
load("../data/selected_mut.Rda")
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
                      ncol = 5) ## facet by gene
```

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)
